# Probability {#probs}

Probability is a measure of the likelihood that an event will occur. Probability is quantified as a number between 0 and 1 (or 0 to 100%), 
where 0 indicates impossibility and 1 indicates certainty, although we will see later that a probability of 0 does not necessarily mean that such an event can never occur. 
The higher the probability of an event, the more likely it is that the event will occur. 

Why is probability important in our field of study (applied health sciences)?

Quantative research methods (often a code name for statistics) use probability theory to make statements about a larger 
population or a data generating process (DGP), as it should be more appropriately called.

In observational studies, we often have to make statements about associations between variables.

In experimental studies (e.g., a randomized controlled trial), we often try to make statements about the effect of an intervention on a 
certain outcome - for instance if a therapy lowers pain by at least 2 points.

Probability theory has its roots in gambling and betting. [Blaise Pascal](https://en.wikipedia.org/wiki/Blaise_Pascal) wrote a letter to [Pierre de Fermat](https://en.wikipedia.org/wiki/Pierre_de_Fermat) in 1654 when a 
French essayist Antoine Gombaud, intrigued by gambling, sought to solve "the problem of points," 
first posed by Luca Paccioli in 1494. The problem asked how to fairly divide the winnings if a game is interrupted before 
its conclusion. Gombaud approached mathematician Blaise Pascal, who collaborated with Pierre de Fermat. Together, they laid 
the groundwork for modern probability theory. Fermat’s method involved listing all possible outcomes and calculating each player’s 
chance of winning, while Pascal developed a backward induction algorithm to assign probabilities. Their work revolutionized mathematics
and influenced fields like economics and actuarial science.

Philosophically speaking, there are two flavors of probability: Probabilities for events that are repeatable respectively have already happened, and probabilities for events that haven't happened yet.

An example for a repeatable event is getting a 6 when throwing a fair dice. We can do this experiment right now by fetching a dice and throwing it.

An example for the latter is the probability of a patient dying within the next 5 years after a certain diagnosis. 
It is hard to argue that this experiment would be repeatable under (almost) identical conditions since every patient is different whereas the dice are typically much more similar.
Here, we could at least put forward that other similar patients have a certain proportion of dying within 5 years.

There are of course events that have not happend ever before, like the creation of artificial general intelligence ([AGI](https://en.wikipedia.org/wiki/Artificial_general_intelligence)). 
Nevertheless, one can still try assign probabilities [when](https://research.aimultiple.com/artificial-general-intelligence-singularity-timing/) such an event would happen.

## Chapters and sub-chapters

There are two steps to cross-reference any heading:

1. Label the heading: `# Hello world {#nice-label}`. 
    - Leave the label off if you like the automated heading generated based on your heading title: for example, `# Hello world` = `# Hello world {#hello-world}`.
    - To label an un-numbered heading, use: `# Hello world {-#nice-label}` or `{# Hello world .unnumbered}`.

1. Next, 

## Captioned figures and tables

Figures and tables *with captions* can also be cross-referenced from elsewhere in your book using `\@ref(fig:chunk-label)` and `\@ref(tab:chunk-label)`, respectively.

See Figure \@ref(fig:nice-fig).

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center', fig.alt='Plot with connected points showing that vapor pressure of mercury increases exponentially as temperature increases.'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Don't miss Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(pressure, 10), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```
