# Probability {#probs}

Probability is a measure of the likelihood that an event will occur. Probability is quantified as a number between 0 and 1, 
where 0 indicates impossibility and 1 indicates certainty, although we will see later that a probability of 0 does not necessarily mean that such an event can never occur. 
The higher the probability of an event, the more likely it is that the event will occur. 

Probability theory started in gambling and betting in a letter written by Blaise Pascal to Pierre de Fermat in 1654 when a 
French essayist Antoine Gombaud, intrigued by gambling, sought to solve "the problem of points," 
first posed by Luca Paccioli in 1494. The problem asked how to fairly divide the winnings if a game is interrupted before 
its conclusion. Gombaud approached mathematician Blaise Pascal, who collaborated with Pierre de Fermat. Together, they laid 
the groundwork for modern probability theory. Fermat’s method involved listing all possible outcomes and calculating each player’s 
chance of winning, while Pascal developed a backward induction algorithm to assign probabilities. Their work revolutionized mathematics
and influenced fields like economics and actuarial science.

## Chapters and sub-chapters

There are two steps to cross-reference any heading:

1. Label the heading: `# Hello world {#nice-label}`. 
    - Leave the label off if you like the automated heading generated based on your heading title: for example, `# Hello world` = `# Hello world {#hello-world}`.
    - To label an un-numbered heading, use: `# Hello world {-#nice-label}` or `{# Hello world .unnumbered}`.

1. Next, 

## Captioned figures and tables

Figures and tables *with captions* can also be cross-referenced from elsewhere in your book using `\@ref(fig:chunk-label)` and `\@ref(tab:chunk-label)`, respectively.

See Figure \@ref(fig:nice-fig).

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center', fig.alt='Plot with connected points showing that vapor pressure of mercury increases exponentially as temperature increases.'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Don't miss Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(pressure, 10), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```
