# Bayes statistics

To gain a richer understanding, I can recommend (optionally) reading the introductory chapters in 
John Kruschke's book [Doing Bayesian Data Analysis](https://nyu-cdsc.github.io/learningr/assets/kruschke_bayesian_in_R.pdf).

As stated in the first chapter, Bayesian statistics is based on the idea that 
[probability](https://en.wikipedia.org/wiki/Bayesian_probability) is a measure of our uncertainty about an event or a parameter.
Here, we use *prior* (i.e., before/outside of our experiment) knowledge 
about a parameter and update this knowledge with new data using the famous Bayes' theorem:

$$
p(\theta | \text{data}) = \frac{p(\text{data} | \theta) \cdot p(\theta)}{p(\text{data})},
$$

where:

- \( p(\theta | \text{data}) \) is the **posterior probability**: the updated probability of the parameter \( \theta \) given the observed data.

- \( p(\text{data} | \theta) \) is the **likelihood**: the probability of observing the data given a certain value of the parameter \( \theta \).

- \( p(\theta) \) is the **prior probability**: the initial belief about the parameter \( \theta \) before seeing the data.

- \( p(\text{data}) \) is the **marginal likelihood** or **evidence**: the probability of observing the data under all possible parameter values.

In our context, $\theta$ is an effect size, a group difference ($\mu_1 - \mu_2$), a correlation ($\rho$), 
a regression coefficient ($\beta$), or a proportion (p)...

## Derivation of Bayes' theorem

We have already come across Bayes' theorem in the context of conditional probability where we have calulated 
the probability of a person having a disease *given* that they have tested positive for it: 

$$
\mathbb{P}(Dpos|Tpos) = \frac{\mathbb{P}(Tpos|Dpos) \cdot \mathbb{P}(Dpos)}{\mathbb{P}(Tpos)}.
$$

The numerator is by definition of the conditional probability just: $\mathbb{P}(Tpos \cap Dpos)$. Let's put this in:

$$
\mathbb{P}(Dpos|Tpos) = \frac{\mathbb{P}(Tpos \cap Dpos)}{\mathbb{P}(Tpos)}.
$$

which is also true by the definition of conditional probability. One can rewrite the denominator as:

$$
\mathbb{P}(Tpos) = \mathbb{P}(Tpos|Dpos)\mathbb{P}(Dpos) + \mathbb{P}(Tpos|Dneg)\mathbb{P}(Dneg) = TP + FP,
$$
since one can have a positive test result if one has the disease (true positive) or if one does not have the disease (false positive). 
This is the so-called [law of total probability](https://en.wikipedia.org/wiki/Law_of_total_probability).
To see this, just draw a binary tree starting with disease status (pos/neg) and test result (pos/neg) as branches, as we did before. 

Summarized, we have:

$$
\mathbb{P}(Dpos|Tpos) = \frac{\mathbb{P}(Tpos|Dpos) \cdot 
\mathbb{P}(Dpos)}{\mathbb{P}(Tpos|Dpos) \cdot \mathbb{P}(Dpos) + \mathbb{P}(Tpos|Dneg) \cdot \mathbb{P}(Dneg)}.
$$

We have proven Bayes theorem for the case of a binary test.

Note that simple fact that if $\mathbb{P}(Dpos)$ (which is the prevalence of the disease)
is small, the probability of having the disease given a positive test result is also small.
In fact, it is arbitrarily small the smaller the prevalence is.
For $\mathbb{P}(Dpos)=0$, we have $\mathbb{P}(Dpos|Tpos) = \frac{0}{0 + \mathbb{P}(Tpos|Dneg) \cdot 1} = 0$, 
assuming that there can still be false positive test results: $\mathbb{P}(Tpos|Dneg) > 0$.

We could easily image that there are not only two states reported by a test. Maybe, it is a more sophisticated test reporting 3 or 4 states.

We would just extend the denominator accordingly 
(e.g., for a 3-state test with labels $T_1, T_2, T_3$, where $T_3$ indicates a high concentration of some component):

$$
\mathbb{P}(Dpos|T_3) = \frac{\mathbb{P}(T_3|Dpos) \cdot 
\mathbb{P}(Dpos)}{\mathbb{P}(T_3|Dpos) \cdot \mathbb{P}(Dpos) + \mathbb{P}(T_3|Dneg) \cdot \mathbb{P}(Dneg)},
$$
which would be the probability that the person has the disease given that the test result is $T_3$.

In this context, we are interested in the probability of a state of disease. 
We estimate this probability based on the test result ($Tpos$ or $Tneg$) and *prior* knowledge about the disease ($\mathbb{P}(Dpos)$).
The prior knowledge does not necessarily have to be known *before* the test, **the point is to combine knowledge in a coherent way**. We will
later see that the order of this combination is not important.

## Bayes' theorem in the context of parameter estimation
If we use the coin flip example, you can always think of probability of a therapy working for instance. 
It is a placeholder for a parameter of interest.

A probability distribution outlines all possible outcomes of a random process along with their associated probabilities. 
For instance, in the case of a coin toss, the distribution is straightforward: there are two outcomes, 
heads and tails, with corresponding probabilities $\theta$ and $1-\theta$. 
For more complex scenarios, such as measuring the height of a randomly chosen individual, 
the distribution is less straightforward. Here, each possible height, say 172 cm or 181 cm, 
is assigned a probability as we have discussed in the context of the normal distribution and continuous distributions
in general.

## Examples

We have jumped into the deep end right away with exercise 2 in the previous chapter. Let's now look at some more examples to get a feeling for 
how the Bayes theorem works.

### Example 1 - defective products {#example_defective_products}

(Example from Script “Statistik und Wahrscheinlichkeitstheorie using R”, 
S.331 ff, Werner Gurker)

A manufacturer claims that the defect rate of their products is only 5%, while the customer believes it to be 10%. 
**Before** the result of a sample inspection is known, we assign both rates an equal 50-50 chance:

\[
\pi(0.05) = \pi(0.10) = 0.5
\]

Assume that in a sample of size 20, there are 3 defective items. 
Using the binomial distribution \(B(20, \theta)\), the data information is given as follows:

\[
p(3|\theta = 0.05) = \binom{20}{3} (0.05)^3 (0.95)^{17} = 0.0596
\]

\[
p(3|\theta = 0.10) = \binom{20}{3} (0.10)^3 (0.90)^{17} = 0.1901
\]

The marginal distribution of \(X\) (the number of defective items in the sample) for \(x = 3\) is as follows:

\[
m(3) = p(3|0.05)\pi(0.05) + p(3|0.10)\pi(0.10) = 0.1249
\]

Here, we consider all parameters in the parameter space of interest. We are only interested in $\theta = 0.05$
vs. $\theta = 0.10$ in this case.

\[
\pi(0.05 \mid X = 3) = \frac{p(3 \mid 0.05) \pi(0.05)}{m(3)} = 0.2387
\]

\[
\pi(0.10 \mid X = 3) = \frac{p(3 \mid 0.10) \pi(0.10)}{m(3)} = \mathbf{0.7613}
\]

A priori, we had no preference for either of the two defect rates. After observing a relatively high 
defect rate of \(3/20 = 15\%\) in the sample, the posterior probability for \(\theta = 0.10\) is 
approximately three times as likely as \(\theta = 0.05\).

Note that in this example the a priori probabilities were equal:

```{r}
# Load ggplot2 package
library(ggplot2)

# Create the data
data <- data.frame(
  theta = factor(c("theta[1] == 0.05", "theta[2] == 0.10"),
                 levels = c("theta[1] == 0.05", "theta[2] == 0.10")),
  probability = c(0.5, 0.5)
)

# Create the bar plot
ggplot(data, aes(x = theta, y = probability, fill = theta)) +
  geom_bar(stat = "identity", width = 0.5) +
  scale_x_discrete(labels = c(expression(theta[1] == 0.05),
                              expression(theta[2] == 0.10))) +
  scale_y_continuous(name = "Probability", limits = c(0, 1)) +
  labs(x = NULL, y = "Probability") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 12, color = "black"),
    axis.title.y = element_text(size = 14, face = "bold")
  )
```

**After* collecting data, we have updated our prior beliefs about the defect rate:

```{r}
# Load ggplot2 package
library(ggplot2)

# Create the data
data <- data.frame(
  theta = factor(c("theta[1] == 0.05", "theta[2] == 0.10"),
                 levels = c("theta[1] == 0.05", "theta[2] == 0.10")),
  probability = c(0.2387, 0.7613)
)

# Create the bar plot
ggplot(data, aes(x = theta, y = probability, fill = theta)) +
  geom_bar(stat = "identity", width = 0.5) +
  scale_x_discrete(labels = c(expression(theta[1] == 0.05),
                              expression(theta[2] == 0.10))) +
  scale_y_continuous(name = "Probability", limits = c(0, 1)) +
  labs(x = NULL, y = "Probability") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 12, color = "black"),
    axis.title.y = element_text(size = 14, face = "bold")
  )
```

Let's assume, we know the manufacturer to be rather untrustworthy. Many inspections in the past
revealed higher defect rates than claimed. For simplicity, we still want to decide between the two 
defect rates. $0.05$ and $0.10$.

How does the calculation change if we assign a much higher probability to the defect rate of $0.10$?
See 




### Example 2 - coin flip


# Exercises

Difficulty levels of exercises: 
**E**: easy, 
**M**: intermediate, 
**D**: difficult

## [E] Exercise 1 - defective product rate
Let's revisit [Example 1](#example_defective_products) and change the prior probabilites to
$\pi(0.05) = 0.1$ and $\pi(0.10) = 0.9$.

Calculate and plot the posterior probabilities for $\theta = 0.05$ and $\theta = 0.10$.
