# Bayes statistics

To gain a richer understanding, I can recommend (optionally) reading the introductory chapters in 
John Kruschke's book [Doing Bayesian Data Analysis](https://nyu-cdsc.github.io/learningr/assets/kruschke_bayesian_in_R.pdf).

As stated in the first chapter, Bayesian statistics is based on the idea that 
[probability](https://en.wikipedia.org/wiki/Bayesian_probability) is a measure of our uncertainty about an event or a parameter.
Here, we use *prior* (i.e., before/outside of our experiment) knowledge 
about a parameter and update this knowledge with new data using the famous Bayes' theorem:

$$
p(\theta | \text{data}) = \frac{p(\text{data} | \theta) \cdot p(\theta)}{p(\text{data})},
$$

where:

- \( p(\theta | \text{data}) \) is the **posterior probability**: the updated probability of the parameter \( \theta \) given the observed data.

- \( p(\text{data} | \theta) \) is the **likelihood**: the probability of observing the data given a certain value of the parameter \( \theta \).

- \( p(\theta) \) is the **prior probability**: the initial belief about the parameter \( \theta \) before seeing the data.

- \( p(\text{data}) \) is the **marginal likelihood** or **evidence**: the probability of observing the data under all possible parameter values.

In our context, $\theta$ is an effect size, a group difference ($\mu_1 - \mu_2$), a correlation ($\rho$), 
a regression coefficient ($\beta$), or a proportion (p)...

## Derivation of Bayes' theorem

We have already come across Bayes' theorem in the context of conditional probability where we have calulated 
the probability of a person having a disease *given* that they have tested positive for it: 

$$
\mathbb{P}(Dpos|Tpos) = \frac{\mathbb{P}(Tpos|Dpos) \cdot \mathbb{P}(Dpos)}{\mathbb{P}(Tpos)}.
$$

The numerator is by definition of the conditional probability just: $\mathbb{P}(Tpos \cap Dpos)$. Let's put this in:

$$
\mathbb{P}(Dpos|Tpos) = \frac{\mathbb{P}(Tpos \cap Dpos)}{\mathbb{P}(Tpos)}.
$$

which is also true by the definition of conditional probability. One can rewrite the denominator as:

$$
\mathbb{P}(Tpos) = \mathbb{P}(Tpos|Dpos)\mathbb{P}(Dpos) + \mathbb{P}(Tpos|Dneg)\mathbb{P}(Dneg) = TP + FP,
$$
since one can have a positive test result if one has the disease (true positive) or if one does not have the disease (false positive). 
This is the so-called [law of total probability](https://en.wikipedia.org/wiki/Law_of_total_probability).
To see this, just draw a binary tree starting with disease status (pos/neg) and test result (pos/neg) as branches, as we did before. 

Summarized, we have:

$$
\mathbb{P}(Dpos|Tpos) = \frac{\mathbb{P}(Tpos|Dpos) \cdot 
\mathbb{P}(Dpos)}{\mathbb{P}(Tpos|Dpos) \cdot \mathbb{P}(Dpos) + \mathbb{P}(Tpos|Dneg) \cdot \mathbb{P}(Dneg)}.
$$

We have proven Bayes theorem for the case of a binary test.

Note that simple fact that if $\mathbb{P}(Dpos)$ (which is the prevalence of the disease)
is small, the probability of having the disease given a positive test result is also small.
In fact, it is arbitrarily small the smaller the prevalence is.
For $\mathbb{P}(Dpos)=0$, we have $\mathbb{P}(Dpos|Tpos) = \frac{0}{0 + \mathbb{P}(Tpos|Dneg) \cdot 1} = 0$, 
assuming that there can still be false positive test results: $\mathbb{P}(Tpos|Dneg) > 0$.

We could easily image that there are not only two states reported by a test. Maybe, it is a more sophisticated test reporting 3 or 4 states.

We would just extend the denominator accordingly 
(e.g., for a 3-state test with labels $T_1, T_2, T_3$, where $T_3$ indicates a high concentration of some component):

$$
\mathbb{P}(Dpos|T_3) = \frac{\mathbb{P}(T_3|Dpos) \cdot 
\mathbb{P}(Dpos)}{\mathbb{P}(T_3|Dpos) \cdot \mathbb{P}(Dpos) + \mathbb{P}(T_3|Dneg) \cdot \mathbb{P}(Dneg)},
$$
which would be the probability that the person has the disease given that the test result is $T_3$.

In this context, we are interested in the probability of a state of disease. 
We estimate this probability based on the test result ($Tpos$ or $Tneg$) and *prior* knowledge about the disease ($\mathbb{P}(Dpos)$).
The prior knowledge does not necessarily have to be known *before* the test, **the point is to combine knowledge in a coherent way**. We will
later see that the order of this combination is not important.

## Bayes' theorem in the context of parameter estimation
If we use the coin flip example, you can always think of probability of a therapy working for instance. 
It is a placeholder for a parameter of interest.

A probability distribution outlines all possible outcomes of a random process along with their associated probabilities. 
For instance, in the case of a coin toss, the distribution is straightforward: there are two outcomes, 
heads and tails, with corresponding probabilities $\theta$ and $1-\theta$. 
For more complex scenarios, such as measuring the height of a randomly chosen individual, 
the distribution is less straightforward. Here, each possible height, say 172 cm or 181 cm, 
is assigned a probability as we have discussed in the context of the normal distribution and continuous distributions
in general.

## Examples

We have jumped into the deep end right away with exercise 2 in the previous chapter. Let's now look at some more examples to get a feeling for 
how the Bayes theorem works.

### Example 1 - defective products {#example_defective_products}

(Example from Script “Statistik und Wahrscheinlichkeitstheorie using R”, 
S.331 ff, Werner Gurker)

A manufacturer claims that the defect rate of their products is only 5%, while the customer believes it to be 10%. 
**Before** the result of a sample inspection is known, we assign both rates an equal 50-50 chance:

\[
\pi(0.05) = \pi(0.10) = 0.5
\]

Assume that in a sample of size 20, there are 3 defective items. 
Using the binomial distribution \(B(20, \theta)\), the data information is given as follows:

\[
p(3|\theta = 0.05) = \binom{20}{3} (0.05)^3 (0.95)^{17} = 0.0596
\]

\[
p(3|\theta = 0.10) = \binom{20}{3} (0.10)^3 (0.90)^{17} = 0.1901
\]

The marginal distribution of \(X\) (the number of defective items in the sample) for \(x = 3\) is as follows:

\[
m(3) = p(3|0.05)\pi(0.05) + p(3|0.10)\pi(0.10) = 0.1249
\]

Here, we consider all parameters in the parameter space of interest. We are only interested in $\theta = 0.05$
vs. $\theta = 0.10$ in this case.

\[
\pi(0.05 \mid X = 3) = \frac{p(3 \mid 0.05) \pi(0.05)}{m(3)} = 0.2387
\]

\[
\pi(0.10 \mid X = 3) = \frac{p(3 \mid 0.10) \pi(0.10)}{m(3)} = \mathbf{0.7613}
\]

A priori, we had no preference for either of the two defect rates. After observing a relatively high 
defect rate of \(3/20 = 15\%\) in the sample, the posterior probability for \(\theta = 0.10\) is 
approximately three times as likely as \(\theta = 0.05\).

Note that in this example the a priori probabilities were equal:

<details open>
<summary>R codes are in the git repository</summary>

```{r bar_plot_code, echo=FALSE}
# Load ggplot2 package
library(ggplot2)

# Create the data
data <- data.frame(
  theta = factor(c("theta[1] == 0.05", "theta[2] == 0.10"),
                 levels = c("theta[1] == 0.05", "theta[2] == 0.10")),
  probability = c(0.5, 0.5)
)
# Create and render the bar plot
ggplot(data, aes(x = theta, y = probability, fill = theta)) +
  geom_bar(stat = "identity", width = 0.5) +
  scale_x_discrete(labels = c(expression(theta[1] == 0.05),
                              expression(theta[2] == 0.10))) +
  scale_y_continuous(name = "Probability", limits = c(0, 1)) +
  labs(x = NULL, y = "Probability") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 12, color = "black"),
    axis.title.y = element_text(size = 14, face = "bold")
  )
```
</details>


**After** collecting data, we have updated our prior beliefs about the defect rate:

<details open>
<summary>R codes are in the git repository</summary>

```{r, echo=FALSE}
# Load ggplot2 package
library(ggplot2)

# Create the data
data <- data.frame(
  theta = factor(c("theta[1] == 0.05", "theta[2] == 0.10"),
                 levels = c("theta[1] == 0.05", "theta[2] == 0.10")),
  probability = c(0.2387, 0.7613)
)

# Create the bar plot
ggplot(data, aes(x = theta, y = probability, fill = theta)) +
  geom_bar(stat = "identity", width = 0.5) +
  scale_x_discrete(labels = c(expression(theta[1] == 0.05),
                              expression(theta[2] == 0.10))) +
  scale_y_continuous(name = "Probability", limits = c(0, 1)) +
  labs(x = NULL, y = "Probability") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    axis.text = element_text(size = 12, color = "black"),
    axis.title.y = element_text(size = 14, face = "bold")
  )
```
</details>

Let's assume, we know the manufacturer to be rather untrustworthy. Many inspections in the past
revealed higher defect rates than claimed. For simplicity, we still want to decide between the two 
defect rates. $0.05$ and $0.10$.

How does the calculation change if we assign a much higher probability to the defect rate of $0.10$?
See [Exercise 1](#exercise_defective_product_rate) for this.


### Example 2 - extending the defective products example

What if we do not want to limit ourselves to the two defect rates of 5% and 10%?
We now get creative and assign the following prior probability **distribution** to the defective rate $\theta$:

<details open>
<summary>R codes are in the git repository</summary>
```{r, echo=FALSE}
# Load ggplot2 package
library(ggplot2)

# Generate beta distribution data
x <- seq(0, 1, length.out = 1000)  # Values between 0 and 1
alpha <- 5  # Shape parameter alpha
beta <- 17  # Shape parameter beta
density <- dbeta(x, shape1 = alpha, shape2 = beta)  # Beta density

# Create a data frame for plotting
data <- data.frame(x = x, density = density)

# Plot the beta density using ggplot2
ggplot(data, aes(x = x, y = density)) +
  geom_line(color = "blue", linewidth = 1.2) +
  labs(
    title = "Beta Distribution Density (Peaks at ~0.2)",
    x = expression(theta),
    y = "Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  ) +
  geom_vline(xintercept = 0.2, linetype = "dashed", color = "red")
```

</details>

If we had to guess, we would say that the defect rate is most likely 20%.
And we are rather sure about this guess. The mode (highest prior probability) of the 
beta distribution is at 0.2. We are also pretty sure that the defect rate is not 0% or above, say, 50%.
This distribution is an expression of our **prior beliefs** about the defect rate.

Now, we **observe 11 defective items in a sample of 20**.
We expect as posterior distribution not a bar plot with two probabilities for the parameter values 
$\theta = 0.05$ and $\theta = 0.10$, but a **distribution** of probabilities for all possible defect rates (in the range of 0-1).
Btw. we could play the same game using a coin of the effectiveness or a therapy as example.

The only thing we need to do is to plug in the information into Bayes' theorem to get the posterior distribution.
The calculations would be complex by hand, so we just use R. We want to understand the concept, so calculating by hand
can be put off until later (it at all necessary for us.)

\[
\textit{p}(\theta \mid X = 3) = \frac{\textit{p}(X = 3 \mid \theta) \cdot \text{Beta}(\theta \mid \alpha, \beta)}{m(3)}
\]

- $\text{Beta}(\theta \mid \alpha, \beta)$ is the prior distribution (density) of the defect rate which captures our beliefs about the defect rate.

- $\mathbb{P}(X = 3 \mid \theta)$ is the likelihood of observing 3 defective items in a sample of 20 given a certain defect rate $\theta$.

- $m(3)$ is the marginal distribution of the data for $X = 3$ considering **all** possible defect rates between 0 and 1.


Let's look at the resulting posterior distribution:

<details open>
<summary>R codes are in the git repository</summary>

```{r, echo=FALSE}
# Load required packages
library(ggplot2)

# Define parameters
alpha_prior <- 5  # Prior alpha (Beta distribution)
beta_prior <- 17  # Prior beta (Beta distribution)
n <- 20           # Sample size
x_obs <- 11       # Number of defective items observed

# Define the likelihood function
likelihood <- function(theta, x, n) {
  choose(n, x) * theta^x * (1 - theta)^(n - x)
}

# Define the prior density (Beta distribution)
prior <- function(theta, alpha, beta) {
  dbeta(theta, shape1 = alpha, shape2 = beta)
}

# Define the unnormalized posterior
unnormalized_posterior <- function(theta, x, n, alpha, beta) {
  likelihood(theta, x, n) * prior(theta, alpha, beta)
}

# Compute the marginal probability using numerical integration
marginal <- integrate(
  f = function(theta) unnormalized_posterior(theta, x_obs, n, alpha_prior, beta_prior),
  lower = 0,
  upper = 1
)$value

# Define the posterior density function
posterior <- function(theta, x, n, alpha, beta, marginal) {
  unnormalized_posterior(theta, x, n, alpha, beta) / marginal
}

# Generate data for plotting the prior and posterior
theta_values <- seq(0, 1, length.out = 1000)
prior_values <- prior(theta_values, alpha_prior, beta_prior)
posterior_values <- posterior(theta_values, x_obs, n,
                              alpha_prior, beta_prior, marginal)

# Create a data frame for plotting
plot_data <- data.frame(
  theta = theta_values,
  Prior = prior_values,
  Posterior = posterior_values
)

# Find the peak (mode) of the posterior distribution
posterior_peak <- theta_values[which.max(posterior_values)]

# Plot the prior and posterior distributions
ggplot(plot_data, aes(x = theta)) +
  geom_line(aes(y = Prior, color = "Prior"),
            linewidth = 1.2, linetype = "dashed") +
  geom_line(aes(y = Posterior, color = "Posterior"), linewidth = 1.2) +
  scale_color_manual(
    values = c("Prior" = "green", "Posterior" = "blue"),
    name = "Distributions"
  ) +
  labs(
    title = "Prior and Posterior Distributions of Theta",
    x = expression(theta),
    y = "Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    legend.position = "top"
  ) + 
  geom_vline(xintercept = 0.2, linetype = "dotted", color = "darkgreen") +
  geom_vline(xintercept = x_obs / n, linetype = "dotted", color = "red") +
  # Annotate the observed defect rate
  annotate(
    "text",
    x = x_obs/n + 0.03,  # Position slightly to the right of the red line
    y = max(posterior_values) * 0.8,  # Adjust height dynamically
    label = paste0("Observed defect rate = ", round(x_obs / n, 2)),
    color = "red",
    size = 4,
    hjust = 0
  ) +
  # Annotate the peak of the posterior
  geom_vline(xintercept = posterior_peak,
             linetype = "dotted", color = "darkblue") +
  annotate(
    "text",
    x = posterior_peak + 0.03,  # Position slightly to the right of the peak
    y = max(posterior_values) * 0.9,  # Adjust height dynamically
    label = paste0("Posterior peak = ", round(posterior_peak, 2)),
    color = "darkblue",
    size = 4,
    hjust = 0
  )
```
</details>

As we can see, the posterior distribution has a new peak at $0.38$. If we had to guess
now, we would probably say that the defect rate is around 38%.
We have updated our beliefs about the defect rate based on the new data.
In the graph, there is also the observed defect rate of 55%. This observed rate (new data)
draws our prior beliefs towards the observed rate. 

- **The more data we observe**, we would observe, the more the posterior distribution would be 
drawn towards the observed rate, because we would be surer due to the large sample size.

- **The stronger our prior beliefs** in a certain value (or range) of $\theta$, the less we are 
convinced by new data.

One really nice aspect of Bayesian statistics is that we get a full probability distribution for 
the parameter of interest, given the prior beliefs and the observed data. 
We then can elegantly make all kinds of statements when looking at the posterior, like the following:

 - The probability that $\theta$ is between $0.30$ and $0.5$ is (area under the posterior) approximately $0.8$:
<details open>
<summary>R codes are in the git repository</summary>
```{r, echo=FALSE}
# Load required libraries
library(ggplot2)

# Define parameters
alpha_prior <- 5  # Prior alpha (Beta distribution)
beta_prior <- 17  # Prior beta (Beta distribution)
n <- 20           # Sample size
x_obs <- 11       # Number of defective items observed

# Define the likelihood function
likelihood <- function(theta, x, n) {
  choose(n, x) * theta^x * (1 - theta)^(n - x)
}

# Define the prior density (Beta distribution)
prior <- function(theta, alpha, beta) {
  dbeta(theta, shape1 = alpha, shape2 = beta)
}

# Define the unnormalized posterior
unnormalized_posterior <- function(theta, x, n, alpha, beta) {
  likelihood(theta, x, n) * prior(theta, alpha, beta)
}

# Compute the marginal probability using numerical integration
marginal <- integrate(
  f = function(theta) unnormalized_posterior(theta, x_obs, n, alpha_prior, beta_prior),
  lower = 0,
  upper = 1
)$value

# Define the normalized posterior density function
posterior_density <- function(theta) {
  unnormalized_posterior(theta, x_obs, n, alpha_prior, beta_prior) / marginal
}

# Numerically integrate the posterior density between 0.3 and 0.5
prob_theta_numerical <- integrate(
  f = posterior_density,
  lower = 0.3,
  upper = 0.5
)$value

# Output the result
#cat("Numerical probability that theta is between 0.3 and 0.5:",
#    prob_theta_numerical, "\n")

# Generate data for the posterior curve
theta_values <- seq(0, 1, length.out = 1000)
posterior_values <- sapply(theta_values, posterior_density)

# Create a data frame for plotting
posterior_data <- data.frame(theta = theta_values, density = posterior_values)

# Plot the posterior distribution with shaded area and annotation
ggplot(posterior_data, aes(x = theta, y = density)) +
  geom_line(color = "blue", linewidth = 1.2) +
  geom_area(
    data = posterior_data[posterior_data$theta >= 0.3 &
                            posterior_data$theta <= 0.5, ],
    aes(x = theta, y = density),
    fill = "orange",
    alpha = 0.6
  ) +
  labs(
    title = "Posterior Distribution of Theta",
    subtitle = "Shaded area shows P(0.3 ≤ Theta ≤ 0.5)",
    x = expression(theta),
    y = "Density"
  ) +
  annotate(
    "text",
    x = 0.4,  # Position of the annotation text along the x-axis
    y = max(posterior_data$density) * 0.5,  # Position along the y-axis
    label = paste0("P(0.3 ≤ Theta ≤ 0.5) = ", round(prob_theta_numerical, 4)),
    color = "black",
    size = 4,
    hjust = 0
  ) +
  theme_minimal(base_size = 14)
```
</details>

- The probability that $\theta$ is below $0.25$ is (area under the posterior):
<details open>
<summary>R codes are in the git repository</summary>
```{r, echo=FALSE}
# Load required libraries
library(ggplot2)

# Define parameters
alpha_prior <- 5  # Prior alpha (Beta distribution)
beta_prior <- 17  # Prior beta (Beta distribution)
n <- 20           # Sample size
x_obs <- 11       # Number of defective items observed

# Define the likelihood function
likelihood <- function(theta, x, n) {
  choose(n, x) * theta^x * (1 - theta)^(n - x)
}

# Define the prior density (Beta distribution)
prior <- function(theta, alpha, beta) {
  dbeta(theta, shape1 = alpha, shape2 = beta)
}

# Define the unnormalized posterior
unnormalized_posterior <- function(theta, x, n, alpha, beta) {
  likelihood(theta, x, n) * prior(theta, alpha, beta)
}

# Compute the marginal probability using numerical integration
marginal <- integrate(
  f = function(theta) unnormalized_posterior(theta, x_obs, n, alpha_prior, beta_prior),
  lower = 0,
  upper = 1
)$value

# Define the normalized posterior density function
posterior_density <- function(theta) {
  unnormalized_posterior(theta, x_obs, n, alpha_prior, beta_prior) / marginal
}

# Numerically integrate the posterior density between 0.0 and 0.25
prob_theta_numerical <- integrate(
  f = posterior_density,
  lower = 0.0,
  upper = 0.25
)$value

# Output the result
#cat("Numerical probability that theta is between 0.0 and 0.25:",
#    prob_theta_numerical, "\n")

# Generate data for the posterior curve
theta_values <- seq(0, 1, length.out = 1000)
posterior_values <- sapply(theta_values, posterior_density)

# Create a data frame for plotting
posterior_data <- data.frame(theta = theta_values, density = posterior_values)

# Plot the posterior distribution with shaded area and annotation
ggplot(posterior_data, aes(x = theta, y = density)) +
  geom_line(color = "blue", linewidth = 1.2) +
  geom_area(
    data = posterior_data[posterior_data$theta <= 0.25, ],
    aes(x = theta, y = density),
    fill = "orange",
    alpha = 0.6
  ) +
  labs(
    title = "Posterior Distribution of Theta",
    subtitle = "Shaded area shows P(0.0 ≤ Theta ≤ 0.25)",
    x = expression(theta),
    y = "Density"
  ) +
  annotate(
    "text",
    x = 0.15,  # Position of the annotation text along the x-axis
    y = max(posterior_data$density) * 0.4,  # Position along the y-axis
    label = paste0("P(0.0 ≤ Theta ≤ 0.25) = ", round(prob_theta_numerical, 4)),
    color = "black",
    size = 4,
    hjust = 0
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", 
                              hjust = 0.5, size = 16),  # Center the title
    plot.subtitle = element_text(hjust = 0.5)  # Center the subtitle
  )
```
</details>

The previous two examples showcase how to **estimate a proportion** using prior knowledge 
and new data.
The fact, that we get a full distribution of the parameter of interest, 
is a key feature of Bayesian statistics and (as far as I know) not available
in frequentist statistics and Null Hypothesis Significance Testing (NHST), which we
will discuss in the next chapter.

## Highest Density Intervals (HDI)

Apart from taking the peak of the posterior distribution, 
another way to **summarize a posterior distribution** is by using 
the **highest density interval (HDI)**. 
The HDI identifies the most credible range of values in a distribution while 
covering a specified portion of the distribution, such as 93%. 
The key feature of the HDI is that every point 
inside the interval has a higher probability density than any point outside 
the interval, making it an effective summary of the most plausible values.

Here is an example for a 93% HDI for the posterior distribution of $\theta$: 

<details open>
<summary>R codes are in the git repository</summary>
```{r, echo=FALSE}
# Load required libraries
library(ggplot2)

# Define parameters
alpha_prior <- 5  # Prior alpha (Beta distribution)
beta_prior <- 17  # Prior beta (Beta distribution)
n <- 20           # Sample size
x_obs <- 11       # Number of defective items observed

# Define the likelihood function
likelihood <- function(theta, x, n) {
  choose(n, x) * theta^x * (1 - theta)^(n - x)
}

# Define the prior density (Beta distribution)
prior <- function(theta, alpha, beta) {
  dbeta(theta, shape1 = alpha, shape2 = beta)
}

# Define the unnormalized posterior
unnormalized_posterior <- function(theta, x, n, alpha, beta) {
  likelihood(theta, x, n) * prior(theta, alpha, beta)
}

# Compute the marginal probability using numerical integration
marginal <- integrate(
  f = function(theta) unnormalized_posterior(theta, x_obs, n, alpha_prior, beta_prior),
  lower = 0,
  upper = 1
)$value

# Define the normalized posterior density function
posterior_density <- function(theta) {
  unnormalized_posterior(theta, x_obs, n, alpha_prior, beta_prior) / marginal
}

# Generate data for the posterior density
theta_values <- seq(0, 1, length.out = 1000)
posterior_values <- sapply(theta_values, posterior_density)

# Create a data frame for the posterior
posterior_data <- data.frame(theta = theta_values, density = posterior_values)

# Compute the 93% HDI
compute_hdi <- function(theta, density, cred_mass = 0.93) {
  sorted_indices <- order(-density)  # Sort by descending density
  cumulative_mass <- cumsum(density[sorted_indices]) / sum(density)  # Cumulative mass
  included <- sorted_indices[cumulative_mass <= cred_mass]  # Indices within HDI
  hdi_range <- range(theta[included])  # Range of HDI
  return(hdi_range)
}

hdi <- compute_hdi(posterior_data$theta,
                   posterior_data$density, cred_mass = 0.93)

# Plot the posterior distribution with the HDI
ggplot(posterior_data, aes(x = theta, y = density)) +
  geom_line(color = "blue", linewidth = 1.2) +
  geom_area(
    data = posterior_data[posterior_data$theta >= hdi[1] &
                            posterior_data$theta <= hdi[2], ],
    aes(x = theta, y = density),
    fill = "orange",
    alpha = 0.6
  ) +
  labs(
    title = "Posterior Distribution with 93% HDI",
    subtitle = paste0("93% HDI: [", round(hdi[1], 4),
                      ", ", round(hdi[2], 4), "]"),
    x = expression(theta),
    y = "Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5)
  ) +
  geom_vline(xintercept = hdi[1], linetype = "dashed", color = "darkred") +
  geom_vline(xintercept = hdi[2], linetype = "dashed", color = "darkred")
```
</details>

**Interpretation**: The 93% HDI for the posterior distribution of $\theta$ is $[0.25, 0.55]$.
With a probability of 93%, the defect rate is between 25% and 55%, given our prior 
beliefs and the observed data.

Note that the HDI does not necessarily have to be symmetric around the peak 
of the distribution. 

There could be, for instance, two peaks in the distribution, as you can see
[here](https://bookdown.org/content/3686/04_files/figure-html/unnamed-chunk-30-1.png) for instance.
In that case, the HDI would consist of two intervals. One could readily image
a real life case for such a distribution. Think of a group of people where the
measurement of interest is the 100 m sprint time. There could be two groups of people: 
The hobby runners and the professional athletes. The distribution of the sprint times would
probably bimodal showing two peaks for the two groups.

There are other ways, one can construct credible intervals.
We could also use the quantiles of the posterior distribution to construct a credible interval.
If the distribution is symmetric, the quantiles are symmetric around the peak of the distribution
and the HDI is the same as the quantile-based credible interval.

A [credible interval](https://en.wikipedia.org/wiki/Credible_interval) 
is **not** the same as a 
[confidence interval](https://en.wikipedia.org/wiki/Confidence_interval) in frequentist statistics.
We will discuss this in the next chapter. 

Briefly, a **credible interval** is a range of values
for a parameter of interest that has a specified probability of containing the 
unobserved parameter.

A **confidence interval** an interval which is expected to typically - when 
constructed **repeatedly** everytime a new sample is drawn - contain the **true, 
but unknown** parameter of interest.

## Exercises

Difficulty levels of exercises: 
**E**: easy, 
**M**: intermediate, 
**D**: difficult

## [E] Exercise 1 - defective product rate {#exercise_defective_product_rate}
Let's revisit [Example 1](#example_defective_products) and change the prior probabilites to
$\pi(0.05) = 0.1$ and $\pi(0.10) = 0.9$.

Calculate and plot the posterior probabilities for $\theta = 0.05$ and $\theta = 0.10$.



## Solutions
Solutions for this chapter can be found [here](https://github.com/jdegenfellner/Script_QM1_ZHAW/tree/main/Solutions_to_Exercises/4_Bayes_Statistics).