# Null Hypothesis Significance Testing (NHST)

Null Hypothesis Significance Testing (NHST) is a statistical method widely used in research, 
including health sciences, to evaluate whether observed data provide sufficient evidence **to refute** a specific hypothesis. 
It operates within a framework of probability and decision-making to address the following question:

Are the observed results likely to occur by chance alone if the null hypothesis is true?
If one decides that the observed results are unlikely to occur by chance alone, 
the null hypothesis is rejected in favor of an alternative hypothesis.

NHST is **not designed to "prove" hypotheses** but rather to provide evidence against the null hypothesis.
Underlying NHST is the idea of [falsifiability](https://en.wikipedia.org/wiki/Falsifiability). 
Sometimes one counter example is enough to reject a hypothesis like "all swans are white". 
Seeing one [black swan](https://en.wikipedia.org/wiki/Falsifiability#/media/File:Black_Swans.jpg), 
proves the hypothesis wrong.

**Key concepts**:

- **Null Hypothesis $H_0$**: This represents the assumption of some specific effect or no effect at all. 
For example, in a clinical trial comparing two treatments, might state that both treatments have the same effect. 
The alternative ist that one treatment is superior to the other.

- **Alternative Hypothesis ($H_1$)**: This is the opposing claim to $H_0$, the logical complement. 
Example: 
   - $H_0: \theta \le 0.4$
   - $H_1: \theta > 0.4$
   - $H_0$ states that the treatment effect is less than or equal to 0.4, while $H_1$ states that the population proportion is greater than 0.4. 
	
- **$p$-value**: The p-value quantifies the probability of obtaining results as extreme as (or more extreme than) 
the observed data, assuming that $H_0$ is true. 
A smaller $p$-value indicates stronger evidence against $H_0$. There are many [misconeptions about p-values](https://www.sciencedirect.com/science/article/abs/pii/S0037196308000620).

- **Significance level ($\alpha$)**: Researchers set a threshold to determine whether to reject $H_1$. 
If the $p$-value is smaller than $\alpha$, $H_0$ is rejected in favor of $H_1$.
Note, that **there is absolutely now special reason to use $\alpha = 0.05$ as a default value**.
To quote [Ronald Fisher](https://link.springer.com/chapter/10.1007/978-1-4612-4380-9_6):
"..., and it is convenient to take twice the standard error as the limit of 
significance; this is roughly equivalent to the corresponding limit P=0.05, ..."


**How NHST works**:

- Formulate Hypotheses: Define $H_0$ (e.g., “The new therapy has no effect”) 
  and $H_1$ (e.g., “The new therapy improves outcomes”).

- Collect data: Perform an experiment or study to gather relevant data.

- Calculate the test statistic: Compute a value based on the sample data that 
  reflects the difference or effect under investigation.
- Compute the p-Value: Determine the probability of observing the 
  test statistic (or more extreme values) if $H_0$ is true.
- Make a Decision: Compare the p-value to the significance level:
  - If $p < \alpha$: Reject $H_0$; evidence suggests $H_1$ is true.
  - If $p \ge \alpha$: Fail to reject $H_0$; insufficient evidence to support $H_1$.

**(Some) Limitations of NHST**: 

- **Focus on p-values with hard cut-offs**: Solely relying on p-values can lead to overinterpretation of
  results without considering practical significance.
	
-	**Dichotomous thinking**: The decision to “reject” or “fail to reject” $H_0$ 
  oversimplifies the complexity of real-world data. This binary thinking
  incentivizes researchers to [focus on statistical significance](https://www.tandfonline.com/doi/full/10.1080/00031305.2018.1447512) 
  rather than real relevance. It happened to me not only once that a colleague 
  looked at me with a sad facial expression announcing that the p-value was "not significant".
  This should be not an issue at all. The focus should be on doing useful analyses in the most
  rigorous way possible.
	
- **Sample size influence**: Large samples can make small, clinically irrelevant 
  differences statistically significant. Example: Given an arbitrarily small difference between means of two groups. There is 
  always a sample size that makes the difference "significant". See [Exercise 2](#exercise2_nhst).

- **[Publication bias](https://en.wikipedia.org/wiki/Publication_bias)**: Many journals tended to publish studies with "significant" results. "Not significant" results
  were often not published. This can lead to a distorted view of the literature.

- **[p-hacking](https://de.wikipedia.org/wiki/P-Hacking)**: In the pursuit for "significant" results 
(which results in publications, which results in tenure), it is natural to do everything to get them. 
  This can include data dredging, selective reporting, and other questionable practices.

See [Kruschke](https://jkkweb.sitehost.iu.edu/articles/Kruschke2013JEPG.pdf) for more limitations of NHST.

In practice, NHST should be accompanied by confidence intervals, 
effect size calculations, and a focus on clinical relevance to provide 
a more comprehensive understanding of the results.

This [Review of NHST](https://f1000research.com/articles/4-621/v3) or 
[this overview](https://en.wikipedia.org/wiki/Statistical_hypothesis_test) might be a good entry point.

This [video](https://www.youtube.com/watch?v=FZ2_hzMyJpY&ab_channel=VeryNormal) could also help to understand the basic concepts.



## Example in the literature

$p$-values are omnipresent in the scientific literature. There a rather few papers in our field that do not contain them. 
Here is an example from a recent paper:


## Binomial test

In the first chapter, we invented the 1000-researcher experiment. There were already encountered hypothesis tests in disguise.
If we would assume that the probability of a false positive is 0.04, 
we would "expect" 40 false positives. We asked, what is the probability of observing
137 or more. This is an example of a hypothesis test:

- $H_0: \theta \le 0.04$
- $H_1: \theta > 0.04$

Under $H_0$, what is the probabilty to see the oberved number of 
false positives (in our case, this is the test statistic) or more?

The answer was $p = 5.551115 \cdot 10^{-16}$.

Any reasonable person would say, that this result did not happen by chance alone and 
therefore concluded, the true, but unknown false positive rate $\theta$ is larger than 0.04.

Formally, this is called a **(one-sided) [binomial test](https://en.wikipedia.org/wiki/Binomial_test)**.

Note, that **$H_1$ is the logical complement of $H_0$**.

```{r}
# binomial test
binom.test(137, 1000, p = 0.04, alternative = "greater")
```

The output in R tells us the following:

- data: 137 and 1000 successes oberved
- alternative hypothesis: true probability of success is greater than 0.04, which we assume afterwards.
- $p$-value$ < 2.2e-16$. This is value is smaller than the precision in R (`.Machine').
- 95 percent confidence interval: $0.1194241$ to $1.0000000$. 
The upper limit of $1$ occurs since we have a once-sided test.
- Sample estimates: Estimating the true (but unknown) proportion from the sample would just be: 
$\frac{137}{1000} = 0.137$

**Two-sided test**:

- $H_0: \theta = 0.04$
- $H_1: \theta \ne 0.04$

One could argue that this is bad style, since we should probably know the direction of the effect.

```{r}
binom.test(137, 1000, p = 0.04, alternative = "two.sided")
```

- alternative = "two.sided", this indicates that we are interested in both directions (higher or lower than $0.04$).
With some experience, one would probably not test for lower when seeing the observed number of 137.
- 95 percent confidence interval: $0.1162817$ to $0.1598810$.

**Interpretation of this interval**: When drawing repeated samples, in 95% percent of the samples,
the so constructed interval (which will be different everytime) contains the true but unknown parameter.

Note, that the "Exact binomial test" was used. There where no approximations made. I would recommend always
using [exact tests](https://en.wikipedia.org/wiki/Exact_test) if available, since we are in the 21th century and computers are fast.

Again, the $\alpha$ level of $0.05$ has nothing special (apart from convention) to it. 
We can also use a $\alpha = 0.14$ level. In this case, we construct confidence intervals 
with a 86% confidence level.

```{r}
binom.test(137, 1000, p = 0.04, alternative = "two.sided", conf.level = 0.86)
```

- confidence interval: $0.1211304$ to $0.1542134$. 

With **smaller coverage probability** (86 instead of 95), we get a **narrower interval**.
Trivially, a 100% confidence interval would be $0$ to $1$ and a 0% confidence interval would be $0.04$ to $0.04$ or any other 
specific value assuming that the true parameter can take any value from $0$ to $1$.

See also [Exercise 3](#exercise3_nhst).

**Comparison with Bayesian version** of estimating $\theta$:

- We cannot include a prior distribution for the paramater $\theta$.
- We cannot calculate the posterior distribution of $\theta$. 
Hence, we cannot make statements like "the probability that $\theta$ is larger than 0.04 is 0.9".
- Prior knowledge could probably be included in the form of the null hypothesis stating, for instance, that $\theta \le 0.2$. 
This could be based on previous studies or expert knowledge.

## Proportions test
......

## (classical) t-test
...........

### One sample t-test
.......

### Two sample t-test
.......Example from before...

### Correlation test
.......

## Type 1 and Type 2 errors
.......

## Simulations based approaches

Through the power of modern computers, we can simulate all kinds of hypothesis tests. 
We just assume that the null hypothesis is true and draw samples from the very distribution.
In the old times, computational ressources were scarce and one had to rely on 
[tables](https://digital.library.adelaide.edu.au/server/api/core/bitstreams/21f22d02-4a39-4650-8f09-71434ad02897/content)
that were precalculated. 

[This](https://github.com/jdegenfellner/ZHAW_Teaching/blob/main/Variability_of_Correlation_under_H_0.R)
shows you how the correlation coefficient behaves under the null hypothesis (that there is no correlation; $\rho = 0$).
This is a very useful tool to understand the behavior of a test statistic under the null hypothesis.
Of course, the probability that a correlation coefficient is exactly 0 is 0 since it's a continuous variable,
but in practice we are interested in "indistinguishable from 0" which is a small value. 
In the Bayesian framework from Kruschke, we defined a region of practical equivalence (ROPE) for this purpose.

We can always ask ourselves: What does my test statistic do *if* the null hypothesis is true?
Then try to simulate it. This helps to understand what constitutes a qualitatively different result 
(from the null hypothesis) considering variability in the data.

## Exercises
.........

### Exercise 1 - frequentist confidence interval {#exercise1_nhst}

- Create 1000 random samples from a binomial distribution with $n = 100$ and $p = 0.38$.
- Calculate the 96% confidence interval for each sample.
- How often was the true parameter ($p = 0.38$) contained in the constructed interval?

### Exercise 2 - everything becomes "significant" {#exercise2_nhst}
Setting: two sample t-test. Assume there is a small difference between the means of two groups.
- Show via simulation that with increasing sample size, the p-value becomes smaller and 
smaller and will be "significant" at some point irresective of how small the true mean 
difference is and how small the $\alpha-$ level is.

### Exercise 3 - binomial test {#exercise3_nhst}

- Create a sample from a binomial distribution with $n = 54$ and $p = 0.68$.
- Perform a two-sided binomial test with $H_0: p = 0.5$.
- Calculate the 90% confidence interval for the sample proportion.
- Calculate the p-vlaue for the two-sided test by "hand" (using dbinom/pbinom in R).