# Null Hypothesis Significance Testing (NHST) {#nhst}

Null Hypothesis Significance Testing (NHST) is a statistical method widely used in research, 
including health sciences, to evaluate whether observed data provide sufficient evidence **to refute** a specific hypothesis. 
It operates within a framework of probability and decision-making to address the following question:

**Are the observed results likely to occur by chance alone if the null hypothesis is true?**

If one decides that the observed results are unlikely to occur by chance alone, 
the null hypothesis is rejected in favor of an alternative hypothesis.

NHST is **not designed to "prove" hypotheses** but rather to provide evidence against the null hypothesis.
If one reads in a paper that there was "no association", this typically means the 
$p$-value was larger than the arbitrary threshold of $0.05$. This does *not* imply that
there is no association (Misconception 2 [here](https://sixsigmadsi.com/wp-content/uploads/2020/10/A-Dirty-Dozen-Twelve-P-Value-Misconceptions.pdf)).

Underlying NHST is the idea of [falsifiability](https://en.wikipedia.org/wiki/Falsifiability). 
Sometimes one counter example is enough to reject a hypothesis like "all swans are white". 
Seeing one [black swan](https://en.wikipedia.org/wiki/Falsifiability#/media/File:Black_Swans.jpg), 
proves the hypothesis wrong.

**Key concepts**:

- **Null Hypothesis $H_0$**: This represents the assumption of some specific effect or no effect at all. 
For example, in a clinical trial comparing two treatments, one might state that both treatments have the same effect. 
The alternative ist that one treatment is superior to the other.

- **Alternative Hypothesis $H_1$**: This is the opposing claim to $H_0$, the logical complement. 
Example: 
   - $H_0: \theta \le 0.4$
   - $H_1: \theta > 0.4$
   - $H_0$ states that the treatment effect is less than or equal to 0.4, 
     while $H_1$ states that the treatment effect is greater than 0.4. 

- **$p$-value**: The $p$-value quantifies the probability of obtaining results as extreme as (or more extreme than) 
the observed data, assuming that $H_0$ is true. 
A smaller $p$-value indicates stronger evidence against $H_0$. 
There are many [misconeptions about $p$-values](https://sixsigmadsi.com/wp-content/uploads/2020/10/A-Dirty-Dozen-Twelve-P-Value-Misconceptions.pdf).

- **Significance level $\alpha$**: Researchers set a threshold to determine whether to reject $H_0$. 
If the $p$-value is smaller than $\alpha$, $H_0$ is rejected in favor of $H_1$.
Note, that **there is absolutely now special reason to use $\alpha = 0.05$ as a default value**.
To quote [Ronald Fisher](https://link.springer.com/chapter/10.1007/978-1-4612-4380-9_6):
"..., and it is convenient to take twice the standard error as the limit of 
significance; this is roughly equivalent to the corresponding limit P=0.05, ..."


**How NHST works**:

- Formulate Hypotheses: Define $H_0$ (e.g., “The new therapy has no effect”) 
  and $H_1$ (e.g., “The new therapy improves outcomes”).

- Determine the necessary sample size to find the effect.

- Collect data: Perform an experiment or study to gather relevant data.

- Calculate the test statistic: Compute a value based on the sample data that 
  reflects the difference or effect under investigation.

- Compute the $p$-value: Determine the probability of observing the 
  test statistic (or more extreme values) if $H_0$ is true.

- Make a Decision: Compare the $p$-value to the significance level:
  - If $p < \alpha$: Reject $H_0$; evidence suggests $H_1$ is true.
  - If $p \ge \alpha$: Fail to reject $H_0$; insufficient evidence to support $H_1$.

**(Some) Limitations of NHST**: 

- **Focus on $p$-values with hard cut-offs**: Solely relying on $p$-values can lead to overinterpretation of
  results without considering practical significance.
	
-	**Dichotomous thinking**: The decision to “reject” or “fail to reject” $H_0$ 
  oversimplifies the complexity of real-world data. This binary thinking
  incentivizes researchers to [focus on statistical significance](https://www.tandfonline.com/doi/full/10.1080/00031305.2018.1447512) 
  rather than real relevance. It happened to me not only once that a colleague 
  looked at me with a sad facial expression announcing that the $p$-value was "not significant".
  This should be not an issue at all. The focus should be on doing useful analyses in the most
  rigorous way possible. Note that the [difference between "significant" and "not significant" is
  itself not significant](https://www.tandfonline.com/doi/abs/10.1198/000313006x152649?casa_token=mn4mIBOnoaIAAAAA:RmZ3RZ6prFHNPWte07aNMKidAsQGmqsjPdCQCpXraH8MCmFuftZmXTsLNzeSLFmXdgJm0Xd8dklEKZdPUw).
	
- **Sample size influence**: Large samples can make small, clinically irrelevant 
  differences "statistically significant". Example: Given an arbitrarily small difference between means of two groups. There is 
  always a sample size that makes the difference "significant". See [Exercise 2](#exercise2_nhst).

- **[Publication bias](https://en.wikipedia.org/wiki/Publication_bias)**: Many journals tended to publish studies with "significant" results. "Not significant" results
  were often not published. This leads to a distorted view of the literature. All rigorously
  conducted studies should be published, irrespective of the results.

- **[$p$-hacking](https://de.wikipedia.org/wiki/P-Hacking)**: In the pursuit for "significant" results 
  (which results in publications, which results in tenure), it is natural to do everything to get them. 
  This can include data dredging, selective reporting, and other questionable practices. 
  In my humble opinion, very often, researchers are not to blame but incentives.

See [Kruschke](https://jkkweb.sitehost.iu.edu/articles/Kruschke2013JEPG.pdf) for more limitations of NHST.

In practice, NHST should be accompanied by confidence intervals, 
effect size calculations, and a focus on clinical relevance to provide 
a more comprehensive understanding of the results.

This [Review of NHST](https://f1000research.com/articles/4-621/v3) or 
[this overview](https://en.wikipedia.org/wiki/Statistical_hypothesis_test) might be a good entry point.

This [video](https://www.youtube.com/watch?v=FZ2_hzMyJpY&ab_channel=VeryNormal) could also help to understand the basic concepts.


## Example in the literature

$p$-values are omnipresent in the scientific literature. There rather few papers in our field that do not contain them. 
We **do not use $p$-values in in descriptive tables** 
(see [here](https://epitodate.com/the-balance-test-fallacy-why-you-shouldnt-put-p-values-in-table-1/) and 
[here](https://www.acpjournals.org/doi/10.7326/0003-4819-147-8-200710160-00010-w1)).

[Here](https://www.tandfonline.com/doi/full/10.1080/09593980601023754#d1e100) is an example from the literature.
Table 2 lists studies and one column contains $p$-values. Note, that it is not good practice to present $p$-values 
(if one should use them at all) as dichotomy: $p < 0.05$. This statement does 
not allow to judge the strength of the evidence against the null hypothesis. Both, $p = 0.00000001$ and $p = 0.049$ would satisfy
the inequality.


## Binomial test

In the first chapter, we invented the 1000-researcher experiment. There, we have already encountered hypothesis tests in disguise.
If we would assume that the probability of a false positive is 0.04, 
we would "expect" 40 false positives. We asked, what is the probability of observing
137 or more. This is an example of a hypothesis test:

- $H_0: \theta \le 0.04$
- $H_1: \theta > 0.04$

Under $H_0$, what is the probabilty to see the oberved number (137) of 
false positives (in our case, this is the test statistic) or more?

The answer was $p = 5.551115 \cdot 10^{-16}$.

A reasonable person would say, that this result did not happen by chance alone and 
therefore conclude, the true, but unknown false positive rate $\theta$ is larger than 0.04.

Formally, this is called a **(one-sided) [binomial test](https://en.wikipedia.org/wiki/Binomial_test)**.

Note, that **$H_1$ is the logical complement of $H_0$**.

```{r}
# binomial test
binom.test(137, 1000, p = 0.04, alternative = "greater")
```

The output in R tells us the following:

- data: 137 and 1000 successes oberved
- alternative hypothesis: true probability of success is greater than 0.04, which we assume afterwards.
- $p$-value < $2.2e-16$. This is value is smaller than the precision in R (`.Machine').
- 95 percent confidence interval: $0.1194241$ to $1.0000000$. 
The upper limit of $1$ occurs since we have a one-sided test.
- Sample estimates: Estimating the true (but unknown) proportion from the sample would just be: 
$\frac{137}{1000} = 0.137$

**Two-sided test**:

- $H_0: \theta = 0.04$
- $H_1: \theta \ne 0.04$

One could argue that this is bad style, since we should probably know the direction of the effect.

```{r}
binom.test(137, 1000, p = 0.04, alternative = "two.sided")
```

- alternative = "two.sided", this indicates that we are interested in both directions (higher or lower than $0.04$).
With some experience, one would probably not test for lower when seeing the observed number of 137.
- 95 percent confidence interval: $0.1162817$ to $0.1598810$.

**Interpretation of this interval**: When drawing repeated samples, in 95% percent of the samples,
the so constructed interval (which will be different everytime) contains the true but unknown 
parameter ([Illustration](https://en.wikipedia.org/wiki/Confidence_interval#/media/File:Normal_distribution_50%25_CI_illustration.svg)). 

Note, that the "Exact binomial test" was used. There were no approximations made. I would recommend always
using [exact tests](https://en.wikipedia.org/wiki/Exact_test) if available, since we are in the 21th century and computers are fast.

Again, **the $\alpha$ level of $0.05$ has nothing special (apart from convention) to it**. 
We can also use a $\alpha = 0.14$ level. In this case, we construct confidence intervals 
with a 86% confidence level. We'll discuss the $\alpha$ level [below](#error_types). 
```{r}
binom.test(137, 1000, p = 0.04, alternative = "two.sided", conf.level = 0.86)
```

- confidence interval: $0.1211304$ to $0.1542134$. 

With **smaller coverage probability** (86 instead of 95), we get a **narrower interval**.
Trivially, a 100% confidence interval would be $0$ to $1$ and a 0% confidence interval would be $0.04$ to $0.04$ or any other 
specific value assuming that the true parameter can take any value from $0$ to $1$.

See also [Exercise 3](#exercise3_nhst).

**Comparison with Bayesian version** of estimating $\theta$:

- We cannot include a prior distribution for the paramater $\theta$.
- We cannot calculate the posterior distribution of $\theta$. 
Hence, we cannot make statements like "the probability that $\theta$ is larger than 0.04 is 0.9".
- Prior knowledge could probably be included in the form of the null hypothesis stating, for instance, that $\theta \le 0.2$. 
This could be based on previous studies or expert knowledge.


## Proportions test

If we are interested in comparing two proportions, 
we can use the [proportions test](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/prop.test).

### One sample case
```{r}
set.seed(443)
heads <- rbinom(1, size = 100,
                prob = 0.5) # create a sample with known probability
prop.test(heads, 100,
          conf.level = 0.94, p = 0.5) # continuity correction TRUE by default
prop.test(heads, 100,
          correct = FALSE,
          conf.level = 0.94, p = 0.5) # no continuity correction
```

- correct = TRUE indicates that a correction ([Yates](https://en.wikipedia.org/wiki/Yates%27s_correction_for_continuity)) 
is used to make the test more accurate and consider the fact that the test statistic is in fact discrete.

- 94% confidence interval: $0.3739955$ to $0.5681618$.

- $\chi^2 = 0.25$. This is the value of the test statistic used in the test. 
Under the null hypothesis (in this case $H_0 = 0.5$), 
the test statistic follows a [$\chi^2$ distribution](https://en.wikipedia.org/wiki/Chi-squared_distribution).

See also [Exercise 4](#exercise4_nhst).

### More than one sample {#proportions_test_more_samples}

```{r}
# Data from Fleiss (1981), p. 139.
# H0: The null hypothesis is that the four populations from which
#     the patients were drawn have the same true proportion of smokers.
# H1: The alternative is that this proportion is different in at
#     least one of the populations.

smokers  <- c(83, 90, 129, 70)
patients <- c(86, 93, 136, 82)
prop.test(smokers, patients)
```

- $p$-value $= 0.005585$. One would argue that this test statistic is unlikely to have come about by chance 
alone and reject the null hypothesis, that all proportions are equal.

- X-squared = 12.6, df = 3. The test statistic is distributed according to a $\chi^2$ distribution 
with 3 [degrees of freedom](https://en.wikipedia.org/wiki/Degrees_of_freedom_(statistics)).
See also [exercise 5](#exercise5_nhst). 
Degrees of freedom is the number of values in the final calculation of a statistic that are free to vary.

## (Classical) $t$-test
The [$t$-test](https://en.wikipedia.org/wiki/Student%27s_t-test) is one of the most 
famous classical statistical tests out there. 
Consider these links: [1](https://www.youtube.com/watch?v=pXuFeRCMTAo&ab_channel=VeryNormal)
[2](https://www.youtube.com/watch?v=pLNuDye_tq4&ab_channel=VeryNormal)
as starting point. 

With the $t$-test, we want to answer the question 

- if the true, but unobserved mean of a population is different from a 
specific value ([one sample $t$-test](https://en.wikipedia.org/wiki/Student%27s_t-test#One-sample_t-test)) or

- if the true, but unobserved means of two populations are different from each other ([two sample $t$-test](https://en.wikipedia.org/wiki/Student%27s_t-test#Two-sample_t-tests)).

Conveniently, R has a built-in function for these tests.


### One sample $t$-test
Let's perform a one sample [$t$-test in R](https://www.sthda.com/english/wiki/one-sample-t-test-in-r) 
(and let's ignore the Shapiro-Wilk test in the link):

```{r}
set.seed(1234)
my_data <- data.frame(
  name = paste0(rep("M_", 10), 1:10),
  weight = round(rnorm(10, 20, 2), 1)
)
head(my_data, 10)
summary(my_data$weight)

library(ggpubr)
ggboxplot(my_data$weight,
          ylab = "Weight (g)", xlab = FALSE,
          ggtheme = theme_minimal())

ggqqplot(my_data$weight, ylab = "Men's weight",
         ggtheme = theme_minimal())

t.test(my_data$weight, mu = 25,
       alternative = "two.sided", conf.level = 0.94)

t.test(my_data$weight, mu = 25,
       alternative = "less", conf.level = 0.94)

t.test(my_data$weight, mu = 25,
       alternative = "greater", conf.level = 0.94)
```

Details output........

### Two sample $t$-test
Let's jump right in and use our example from the previous chapter where we performed the Bayesian $t$-test:

```{r, echo=TRUE}
(y1 <- c(-.5, 0, 1.2, 1.2, 1.2, 1.9, 2.4, 3) * 100)
(y2 <- c(-1.2, -1.2, -.5, 0, 0, .5, 1.1, 1.9) * 100)

length(y1)
length(y2)

data <- data.frame(y1, y2)
psych::describe(data)

# Boxplot:
data.frame(y = c(y1, y2), group = c(rep(1, 8), rep(2, 8))) %>%
  ggplot(aes(x = factor(group), y = y)) +  # Use factor for discrete x-axis
  geom_boxplot() +                        # Add boxplot layer
  geom_jitter(width = 0.1)                # Add jitter for individual
# -> Visually, there seems to be a difference between the two groups.

t.test(y1, y2, conf.level = 0.93)
```

- The test statistic is a $t$-distribution with $13.928$ degrees of freedom.
- The value of the test statistic is $2.1867$.
- $H_0: \mu_1 = \mu_2$. The group means are equal.
- $H_1: \mu_1 \ne \mu_2$. The group means are different.
- $p$-value $= 0.04634$. In the classical framework, 
  this would be considered "significant" at the $\alpha = 0.05$ level 
  and one would reject the null hypothesis and accept the alternative hypothesis.
  **In the Baysian framework, we abstained from making a decision**.
- 93% confidence interval for the difference in means is rather wide: 
  $12.55847$ to $232.44153$.

  Let's try to visualize this. 
  
  **Under the assumption** that there is **no difference in means**,
  the [test statistic](https://en.wikipedia.org/wiki/Student%27s_t-test#Equal_or_unequal_sample_sizes,_unequal_variances_(sX1_%3E_2sX2_or_sX2_%3E_2sX1)) 
  
  $$ t = \frac{\bar{X}_1 - \bar{X}_1}{\sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}} $$
  
  would be distributed according to a $t$-distribution with 13.928 degrees of freedom.

```{r}
# Load ggplot2
library(ggplot2)

# Define degrees of freedom
df <- 13.928

# Define the range for x and the critical t-values
x <- seq(-4, 4, length.out = 500)
critical_t <- 2.1867

# Create a data frame with x and corresponding density values
t_dist <- data.frame(
  x = x,
  density = dt(x, df)
)

# Plot the t-distribution with shaded tail areas
ggplot(t_dist, aes(x = x, y = density)) +
  # Add the main t-distribution curve
  geom_line(linewidth = 1, color = "blue") +
  # Add shaded areas below the curve outside the critical t-values
  geom_ribbon(
    data = subset(t_dist, x < -critical_t),
    aes(ymin = 0, ymax = density),
    fill = "red",
    alpha = 0.3
  ) +
  geom_ribbon(
    data = subset(t_dist, x > critical_t),
    aes(ymin = 0, ymax = density),
    fill = "red",
    alpha = 0.3
  ) +
  # Add vertical lines for the critical t-values
  geom_vline(xintercept = c(-critical_t, critical_t),
             linetype = "dashed", color = "black") +
  # Annotate the critical t-values
  annotate("text", x = -critical_t, y = 0.05,
           label = paste0("-t = ", critical_t), angle = 90, vjust = -0.5) +
  annotate("text", x = critical_t, y = 0.05,
           label = paste0("t = ", critical_t), angle = 90, vjust = -0.5) +
  # Add labels and style
  labs(
    title = "t-Distribution with Shaded Critical Areas (Two-Sided Test)",
    subtitle = paste("Degrees of Freedom:", df),
    x = "t",
    y = "Density"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),  # Center the title
    plot.subtitle = element_text(hjust = 0.5)  # Center the subtitle
  )

# p-value manually:
pt(-2.1867, df) * 2
```

The plot shows the $t$-distribution. Marked in red are the areas where the test statistic 
takes the value we observed or "more extreme" values. With "more extreme" we mean values that are
further away from 0 in both directions since we conducted a two-sided test. The area under the curve
is the $p$-value. As you can see, the $p$-value is the sum of the two red areas and matches the output 
of the $t$-test function in R.


## Correlation test

In the chapter about descriptive statistics, we calculated the (Pearson) correlation coefficient 
to measure the strength of the **linear** relationship between two variables.

Often, the null hypothesis for the correlation coefficient is that there is no correlation between 
the two variables ($\rho=0$). One could argue that this is a rather baseless assumption. In reality,
the true correlation coefficient is probably not exactly 0 and one could argue more precisely a range
of plausible values for $\rho$ for the specific case at hand. Often, one can see an ocean of $p$-values
in the literature, where the correlation coefficient is tested against 0. This is superfluous.
For example, if the sample size is $n=234$ and the sample correlation coefficient is $r = 0.76$, it 
is very unlikely that the true correlation coefficient is 0. One does not need a hypothesis test to 
know this (see [exercise 6](#exercise6_nhst)).

We can do a correlation test in R with the `cor.test` function. 
[This article](https://www.sthda.com/english/wiki/correlation-test-between-two-variables-in-r) is also helpful.
Maybe, let's take the part about the Shapiro-Wilk test not too seriously.

The test statistic(s) for the test(s) can be found [here](https://en.wikipedia.org/wiki/Pearson_correlation_coefficient#Inference).
If we really want to take the result of such a test seriously, we need to check the assumptions of the test.

...........

See also [here](https://github.com/jdegenfellner/ZHAW_Teaching/blob/main/Variability_of_Correlation_under_H_0.R) 
to see the variability of the correlation coefficient under the null hypothesis.

## Type 1 and Type 2 errors {#error_types}
```{r error_table_fixed, echo=FALSE}
library(knitr)

# Define the data
data <- data.frame(
  "Decision about null hypothesis (H0)" = c("Not reject", "Reject"),
  "Null hypothesis (H0) is true" = c(
    "Correct inference (true negative)\n(probability = 1 - α)",
    "Type I error (false positive)\n(probability = α)"
  ),
  "Null hypothesis (H0) is false" = c(
    "Type II error (false negative)\n(probability = β)",
    "Correct inference (true positive)\n(probability = 1 - β)"
  )
)

# Render the table
kable(data, format = "html", align = "c", row.names = FALSE,
      col.names = c("Decision about null hypothesis (H0)",
                    "Null hypothesis (H0) is true",
                    "Null hypothesis (H0) is false"))
```

## The frequentist confidence interval

**In frequentist statistics, a [confidence interval (CI)](https://en.wikipedia.org/wiki/Confidence_interval) is an interval which is expected to 
typically (with repeated sampling) contain the true but unknown parameter with a certain frequency (probability).**

It is very important to understand that for every sample, the confidence interval is different and in the long run,
these intervals will contain the true parameter in a certain percentage of cases.

This is a [nice visualization](https://en.wikipedia.org/wiki/Confidence_interval#/media/File:Normal_distribution_50%25_CI_illustration.svg) 
of the concept.
See also [exercise 7](#exercise7_nhst).




## Simulations based approaches

Through the power of modern computers, we can simulate all kinds of hypothesis tests. 
We just assume that the null hypothesis is true and draw samples from the very distribution.
In the old times, computational ressources were scarce and one had to rely on 
[tables](https://digital.library.adelaide.edu.au/server/api/core/bitstreams/21f22d02-4a39-4650-8f09-71434ad02897/content)
that were precalculated. 

[This](https://github.com/jdegenfellner/ZHAW_Teaching/blob/main/Variability_of_Correlation_under_H_0.R)
shows you how the correlation coefficient behaves under the null hypothesis (that there is no correlation; $\rho = 0$).
This is a very useful tool to understand the behavior of a test statistic under the null hypothesis.
Of course, the probability that a correlation coefficient is exactly 0 is 0 since it's a continuous variable,
but in practice we are interested in "indistinguishable from 0" which is a small value. 
In the Bayesian framework from Kruschke, we defined a region of practical equivalence (ROPE) for this purpose.

**We ask ourselves**: What does my test statistic do *if* the null hypothesis is true?
Then try to simulate it. This helps to understand what constitutes a qualitatively different result 
(from the null hypothesis) considering variability in the data.

## Exercises

### Exercise 1 - frequentist confidence interval {#exercise1_nhst}

- Create 1000 random samples from a binomial distribution with $n = 100$ and $p = 0.38$.
- Calculate the 96% confidence interval for each sample using R.
- How often was the true parameter ($p = 0.38$) contained in the constructed interval?

### Exercise 2 - everything becomes "significant" {#exercise2_nhst}

Setting: two sample $t$-test. Assume there is a small difference between the means of two groups.
- Show via simulation that with increasing sample size, the $p$-value becomes smaller and 
smaller and will be "significant" at some point irresespective of how small the true mean 
difference is and how small the $\alpha-$ level is.

### Exercise 3 - binomial test {#exercise3_nhst}

- Create a sample from a binomial distribution with $n = 54$ and $p = 0.68$.
- Perform a two-sided binomial test with $H_0: p = 0.5$.
- Calculate the 90% confidence interval for the sample proportion.
- Calculate the p-value for the two-sided test by "hand" (using dbinom/pbinom in R).

### Exercise 4 - proportions test {#exercise4_nhst}

- Create a sample from a binomial distribution with $n = 100$ and $p = 0.5$.
- Perform a proportions test with $H_0: p = 0.5$ and interpret the results.
- Perform the proportions test with the whole range of possible proportions
  $H_0: p = 0.01 \cdots p = 0.99$ in steps of $0.01$. 
  And plot the p-values on the y-axis and 
  the assumed proportion on the x-axis. 
  This is called a [p-value function](https://www.tandfonline.com/doi/full/10.1080/00031305.2018.1556735#:~:text=The%20p%2Dvalue%20function%20provides,for%20the%20parameter%20of%20interest.).

### Exercise 5 - proportions test 2 {#exercise5_nhst}
- Use the data from the smokers proportions test [example above](#proportions_test_more_samples).
- Draw a $\chi^2$ distribution with 3 degrees of freedom and calculate the 
  probability of observing a value of 12.6 or larger.

### Exercise 6 - correlation coefficent {#exercise6_nhst}
- Create a sample of $234$ pairs of uncorrelated observations $(x_i,y_i)$. 
  $X_i$ and $Y_i$ are drawn from a normal distribution with mean 0 and standard deviation 1.
- Calculate the sample correlation coefficient $r$.
- Repeat this 1000 times.
- How often was the sample correlation coefficient larger than 0.76?

### Exercise 7 - coverage frequency of CI {#exercise7_nhst} 
- Create a sample (vary the sample size, start small) from a normal distribution with mean 0 and standard deviation 1.
- Calculate the 93% confidence interval for the mean.
- Repeat this 1000 times.
- How often was the true mean (0) contained in the constructed interval?